Adaptive Video Streaming with PID-Based Quality Control and Optimized Bitstream Extraction

### 2015-02-06 tmm.tex
introduction with shorter paragraphs; the 2rd version sent to Sun

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with the challenge, modern video streaming systems should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, which allows scalability of video bitrate and quality with only one single bitstream. The generally-known term SVC usually refers to the scalable extension of the H.264/AVC video coding standard \cite{SVCOverview}. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into multiple layers: one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most existing video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity simulcast provides is strictly limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, SVC is considered to be a better choice for adaptive video streaming systems. While the work of this paper can be partly used in a simulcast system, it will mainly focus on the adaptive streaming system using SVC.

In a video streaming system using SVC, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constrain of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

>>>>

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, which allows scalability of video bitrate and quality with only one single bitstream. The generally-known term SVC usually refers to the scalable extension of the H.264/AVC video coding standard \cite{SVCOverview}. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most commercial video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, we choose to adopt SVC in our adaptive video streaming system. While the work of this paper can be partly used in a simulcast system, it will mainly focus on the adaptive streaming system using SVC.

In the SVC video streaming system, one scalable stream with enhancement data packets is produced according to actual requirement and stored at the server for streaming (how to configure and produce the stream is not in the scope of this paper). During the streaming process, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constrain of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.
